---
title: "HEADWAY - a tool for predicting train delays"
author: "Elisabeth Ericson & Adrián León"
date: "December 2021"
output: 
  html_document: 
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, error = F, message = F, 
                      results = F, cache = T, fig.align = "center")
```

```{r libraries & options}

library(tidyverse)
library(sf)
library(lubridate)
library(riem)
library(imputeTS)
library(gridExtra)

options(scipen = 999) # block scientific notation
set.seed(508) # random seed

# set coordinate system
nj_crs = "EPSG:3424" # NAD83/New Jersey
# nj_crs = "EPSG:4326" # WGS84/required for ggmap

# set map styling options
mapTheme <-
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(face = "plain"),
    plot.subtitle = element_text(face = "italic"),
    # plot.caption = element_text(hjust = 0)
  )

# set plot styling options
plotTheme <-
  theme(
    axis.ticks = element_blank(),
    legend.title = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line(color = "gray75", size = 0.1),
    panel.grid.minor = element_line(color = "gray75", size = 0.1),
    plot.title = element_text(face = "plain"),
    plot.subtitle = element_text(face = "italic"),
    # plot.caption = element_text(hjust = 0)
  )

```


## Introduction




## Data exploration

### Data sources

```{r read in data}

# read in train departures
departures_raw <- 
  rbind(read.csv("data/2019_09.csv"), 
        read.csv("data/2019_10.csv"))

# read in station coordinates
stations <- 
  read_csv("data/stops.csv") %>%
  st_as_sf(coords = c("stop_lon", "stop_lat")) %>%
  st_set_crs("EPSG:4326") %>%
  st_transform(nj_crs)

```

### System and line overview

- excluded Amtrak due to lack of schedule data
- (at least initially) selected line(s) by data quality
- NOTE: lines missing sequence data are also the ones missing schedule data, so we do have sequence info for all relevant trains/stops

```{r check data quality by line}

line_stats <-
  filter(departures_raw, type == "NJ Transit") %>%
  group_by(line) %>%
  summarize(
    total_records = n(),
    delay_NA_pct = round(sum(is.na(delay_minutes)) / total_records, 3),
    sequence_NA_pct = round(sum(is.na(stop_sequence)) / total_records, 3),
    from_NA_pct = round(sum(is.na(from_id)) / total_records, 3),
    to_NA_pct = round(sum(is.na(to_id)) / total_records, 3),
    delay_mean = mean(delay_minutes, na.rm = T),
    delay_median = median(delay_minutes, na.rm = T)
  )

system_stats <-
  filter(departures_raw, type == "NJ Transit") %>%
  summarize(
    total_records = n(),
    delay_NA_pct = round(sum(is.na(delay_minutes)) / total_records, 3),
    sequence_NA_pct = round(sum(is.na(stop_sequence)) / total_records, 3),
    from_NA_pct = round(sum(is.na(from_id)) / total_records, 3),
    to_NA_pct = round(sum(is.na(to_id)) / total_records, 3),
    delay_mean = mean(delay_minutes, na.rm = T),
    delay_median = median(delay_minutes, na.rm = T)
  )

ggplot(line_stats) +
  geom_bar(aes(x = reorder(line, -delay_NA_pct), y = delay_NA_pct), stat = "identity") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  geom_hline(yintercept = system_stats$delay_NA_pct, linetype = 3) +
  labs(title = "Missing schedule data by line",
       subtitle = "Line represents system average",
       y = "Percent") +
  theme(axis.title = element_blank()) +
  plotTheme

ggplot(line_stats) +
  geom_bar(aes(x = reorder(line, -delay_mean), y = delay_mean), stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = system_stats$delay_mean, linetype = 3) +
  labs(title = "Mean delay by line",
       subtitle = "Line represents system average",
       y = "Minutes") +
  theme(axis.title.y = element_blank()) +
  plotTheme

ggplot(line_stats) +
  geom_bar(aes(x = reorder(line, -delay_median), y = delay_median), stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = system_stats$delay_median, linetype = 3) +
  labs(title = "Median delay by line",
       subtitle = "Line represents system average",
       y = "Minutes") +
  theme(axis.title.y = element_blank()) +
  plotTheme

ggplot(line_stats) +
  geom_bar(aes(x = reorder(line, -sequence_NA_pct), y = sequence_NA_pct), stat = "identity") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  geom_hline(yintercept = system_stats$delay_NA_pct, linetype = 3) +
  labs(title = "Missing sequence data by line",
       subtitle = "Line represents system average",
       y = "Percent") +
  theme(axis.title = element_blank()) +
  plotTheme


```


```{r train stats}

train_stats <-
  filter(departures_raw, type == "NJ Transit") %>%
  group_by(line, train_id) %>%
  summarize(
    total_records = n(),
    delay_NA_pct = round(sum(is.na(delay_minutes)) / total_records, 3),
    sequence_NA_pct = round(sum(is.na(stop_sequence)) / total_records, 3),
    from_NA_pct = round(sum(is.na(from_id)) / total_records, 3),
    to_NA_pct = round(sum(is.na(to_id)) / total_records, 3),
    delay_mean = mean(delay_minutes, na.rm = T),
    delay_median = median(delay_minutes, na.rm = T)
  )

ggplot(train_stats) +
  geom_bar(aes(x = reorder(train_id, -delay_mean), y = delay_mean), stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = system_stats$delay_mean, linetype = 3) +
  labs(title = "Mean delay by train",
       subtitle = "Line represents system average",
       y = "Minutes") +
  theme(axis.title.y = element_blank()) +
  plotTheme

```


## Feature engineering

### Preparing departure-level data


```{r }

# line_list <- c("Pascack Valley")

departures <-
  dplyr::select(departures_raw, -date, -actual_time, -type) %>%
  filter(!is.na(delay_minutes) & status != "cancelled") %>%
  mutate(
    # standardize existing features
    train_id = as.factor(train_id),
    from_id = as.factor(from_id),
    to_id = as.factor(to_id),
    scheduled_time = ymd_hms(scheduled_time),
    # create line-level station identifier
    station = paste(to, line, sep="_"),
    # create new time features
    week = isoweek(scheduled_time),
    dotw = wday(scheduled_time, label = T),
    hour = hour(scheduled_time),
    interval60 = floor_date(scheduled_time, unit = "hour"),
    interval30 = floor_date(scheduled_time, unit = "30 mins"),
    interval15 = floor_date(scheduled_time, unit = "15 mins")
  )


# departures <-
#   dplyr::select(departures_raw, -date, -actual_time, -type) %>%
#   filter(#line %in% line_list & 
#            status != "cancelled") %>%
#   mutate(
#     # standardize existing features
#     train_id = as.factor(train_id),
#     from_id = as.factor(from_id),
#     to_id = as.factor(to_id),
#     scheduled_time = ymd_hms(scheduled_time),
#     # create line-level station identifier
#     station = paste(to, line, sep="_"),
#     # create new time features
#     week = isoweek(scheduled_time),
#     dotw = wday(scheduled_time, label = T),
#     hour = hour(scheduled_time),
#     interval60 = floor_date(scheduled_time, unit = "hour"),
#     interval30 = floor_date(scheduled_time, unit = "30 mins"),
#     interval15 = floor_date(scheduled_time, unit = "15 mins")
#   )



```


### Accounting for weather

- in addition to temperature, precip in last hr, and wind speed, included visibility + added new wind feature to attempt to capture storms: peak wind
- peak wind = peak wind gust if available; if not, highest regular gust if available; if not, highest recorded wind speed
- also wanted to do lagged weather, but unsure how/if time/etc


```{r weather, fig.height=7}

# load weather data
weather_data <- 
  riem_measures(
    station = "EWR", date_start = "2019-09-01", date_end = "2019-11-01"
  ) %>%
  dplyr::select(
    time = valid,
    temperature = tmpf,
    windspeed = sknt,
    precip_1h = p01i,
    visibility = vsby,
    gust = gust,
    peak_wind_gust
  ) %>%
  mutate(
    interval60 = floor_date(time, unit = "hour"),
    interval15 = floor_date(time, unit = "15 min"),
    peak_wind = case_when(
      peak_wind_gust > 0 ~ peak_wind_gust,
      gust > 0 ~ gust,
      windspeed >= 0 ~ windspeed
    )
  )

# create weather panel and impute missing values
weather_panel <-
  weather_data %>%
  group_by(interval60) %>%
  summarize(
    temperature = mean(temperature, na.rm = T),
    precip_1h = mean(precip_1h, na.rm = T),
    visibility = mean(visibility, na.rm = T),
    windspeed = mean(windspeed, na.rm = T),
    peak_wind = if_else(
      !is.na(max(peak_wind)), max(peak_wind), max(windspeed)
    )
  ) %>%
  mutate(
    temperature_na = if_else(!is.nan(temperature), temperature, NA_real_),
    temperature = na_interpolation(temperature_na),
    precip_1h_na = if_else(!is.nan(precip_1h), precip_1h, NA_real_),
    precip_1h = na_interpolation(precip_1h_na)
  ) %>%
  dplyr::select(-temperature_na, -precip_1h_na)

# create charts by weather indicator
grid.arrange(
  ncol= 1,
  top = "Weather data - Newark Liberty International Airport (EWR) - Sept & Oct 2019",
  ggplot(weather_panel, aes(interval60, precip_1h)) + geom_line() + 
    labs(title = "Precipitation in last hour", x = "Hour", y = "Inches") + plotTheme,
  ggplot(weather_panel, aes(interval60, temperature)) + geom_line() +
    labs(title = "Temperature", x = "Hour", y = "Degrees F") + plotTheme,
  ggplot(weather_panel, aes(interval60, windspeed)) + geom_line() +
    labs(title = "Wind speed", x = "Hour", y = "Knots") + plotTheme,
  ggplot(weather_panel, aes(interval60, peak_wind)) + geom_line() +
    labs(title = "Peak wind", x = "Hour", y = "Knots") + plotTheme,
  ggplot(weather_panel, aes(interval60, visibility)) + geom_line() +
    labs(title = "Visibility", x = "Hour", y = "Miles") + plotTheme
)


```


## Creating the space-time panel

```{r }

# create empty panel of all station-hour combinations
empty_panel <- 
  expand_grid(
    interval60 = unique(departures$interval60),
    station = unique(departures$station)
  ) %>%
  arrange(station, interval60)



# compute delay minutes and trains per station per hour
station_delays <- 
  departures %>%
  group_by(station, interval60) %>%
  summarize(
    trains = n(),
    delay_minutes = sum(delay_minutes)
  )

# create station panel and join weather features
station_panel <-
  left_join(empty_panel, station_delays) %>%
  mutate(
    trains = replace_na(trains, 0),
    delay_minutes = replace_na(delay_minutes, 0),
    week = isoweek(interval60),
    day = wday(interval60, label = T),
    hour = hour(interval60)
  ) %>%
  left_join(weather_panel, by = "interval60") %>%
  arrange(station, interval60) %>%
  mutate(
    lag_1h = dplyr::lag(delay_minutes, 1),
    lag_2h = dplyr::lag(delay_minutes, 2),
    lag_3h = dplyr::lag(delay_minutes, 3),
    lag_4h = dplyr::lag(delay_minutes, 4),
    lag_12h = dplyr::lag(delay_minutes, 12),
    lag_24h = dplyr::lag(delay_minutes, 24),
    lag_1wk = dplyr::lag(delay_minutes, 168),
    lag_2wk = dplyr::lag(delay_minutes, 336),
    lag_3wk = dplyr::lag(delay_minutes, 504),
    lag_4wk = dplyr::lag(delay_minutes, 672)
  ) %>%
  ungroup() 

#%>%
  #filter(interval60 >= as.Date("2019-09-02")) # remove day used to create lag features

```


## Model testing

```{r create training and test sets}


training <- filter(station_panel, week <= 41)
test <- filter(station_panel, week > 41)


```

- mention how panel was created: all unique station/line combos * all hours/days from _weather_ panel


```{r create regression versions}

# --- define regressions to test ---

timeVars <- c("hour", "day")

spaceVars <- c("station")

simpleWeatherVars <- c("temperature", "precip_1h", "windspeed")

weatherVars <- 
  c("temperature", "precip_1h", "visibility", "windspeed", "peak_wind")

lagVarsOG <- c("lag_1h", "lag_2h", "lag_3h", "lag_4h", "lag_12h", "lag_24h")

lagVars <- c("lag_1wk", "lag_2wk")

trafficVars <- c("trains")

# A - time alone
timeModel <- as.formula(
  paste("delay_minutes",
        paste(timeVars, 
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg1 <- lm(timeModel, data = training)

# B - space alone
spaceModel <- as.formula(
  paste("delay_minutes",
        paste(spaceVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg2 <- lm(spaceModel, data = training)
  
# C - space and time
spaceTimeModel <- as.formula(
  paste("delay_minutes",
        paste(spaceVars,
              timeVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg3 <- lm(spaceTimeModel, data = training)

# D - space, time, and lagged delays
lagModel <- as.formula(
  paste("delay_minutes",
        paste(timeVars, 
              spaceVars,
              lagVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg4 <- lm(lagModel, data = training)

# E - space, time, lagged delays, and weather
weatherModel <- as.formula(
  paste("delay_minutes",
        paste(timeVars, 
              spaceVars,
              lagVars,
              weatherVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg5 <- lm(weatherModel, data = training)

# F - space, time, lagged delays, weather, and other scheduled trains
trafficModel <- as.formula(
  paste("delay_minutes",
        paste(timeVars, 
              spaceVars,
              lagVars,
              weatherVars, 
              trafficVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg6 <- lm(trafficModel, data = training)

# G - space, NO TIME, lagged delays, more complex weather, and other scheduled trains
noTimeModel <- as.formula(
  paste("delay_minutes",
        paste(spaceVars,
              weatherVars, 
              lagVars,
              trafficVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg7 <- lm(noTimeModel, data = training)

# H - space, NO TIME, hour+week lagged delays, more complex weather, and other scheduled trains
allLagsModel <- as.formula(
  paste("delay_minutes",
        paste(spaceVars,
              weatherVars, 
              lagVarsOG,
              lagVars,
              trafficVars,
              collapse = " + ", sep = " + "),
        sep = " ~ "))

reg8 <- lm(allLagsModel, data = training)


# --- validate on test set ---



# define function to generate predictions
model_pred <- function(data, fit) {
  predictions <- predict(fit, newdata = data)
}

# nest test set by week
test_week_nest <- 
  test %>%
  nest(-week)

# generate predictions by week
week_predictions <-
  test_week_nest %>%
  mutate(
    A_Time_FE = map(.x = data, fit = reg1, .f = model_pred),
    B_Space_FE = map(.x = data, fit = reg2, .f = model_pred),
    C_SpaceTime_FE = map(.x = data, fit = reg3, .f = model_pred),
    D_SpaceTime_Lag = map(.x = data, fit = reg4, .f = model_pred),
    E_SpaceTime_Lag_Weather = map(.x = data, fit = reg5, .f = model_pred),
    F_SpaceTime_Lag_Weather_Traffic = map(.x = data, fit = reg6, .f = model_pred),
    G_NoTime = map(.x = data, fit = reg7, .f = model_pred),
    H_AllLags = map(.x = data, fit = reg8, .f = model_pred)
  )

# calculate error metrics by week
week_predictions <-
  week_predictions %>%
  gather(Regression, Prediction, -data, -week) %>%
  mutate(
    Observed = map(data, pull, delay_minutes),
    Absolute_Error = map2(Observed, Prediction, ~ abs(.x - .y)),
    MAE = map_dbl(Absolute_Error, mean),
    sd_AE = map_dbl(Absolute_Error, sd)
  )



# predictions_nested <-
#   predictions %>%
#   gather(Regression, Prediction, -data) %>%
#   mutate(
#     Observed = map(data, pull, delay_minutes),
#     Absolute_Error = map2(Observed, Prediction, ~ abs(.x - .y)),
#     MAE = map_dbl(Absolute_Error, mean),
#     sd_AE = map_dbl(Absolute_Error, sd)
#   )
# 
# predictions_nested %>%
#   dplyr::select(Regression, MAE) %>%
#   gather(variable, MAE, -Regression) %>%
#   ggplot(aes(Regression, MAE)) +
#     geom_bar(aes(fill = Regression), stat = "identity", position = "dodge") +
#     labs(
#       title = "Mean absolute errors",
#       subtitle = "by model"
#     ) +
#     plotTheme +
#     theme(legend.position = "bottom")
# 
# predictions <-
#   predictions_nested %>%
#   unnest(cols = everything())
# 
# prediction_panel
#   

```

```{r validation: plot MAE by model and week, fig.height=3.5, fig.width=10}

week_predictions %>%
  dplyr::select(week, Regression, MAE) %>%
  gather(variable, MAE, -Regression, -week) %>%
  ggplot(aes(week, MAE)) +
    geom_bar(aes(fill = Regression), stat = "identity", position = "dodge") +
    # scale_fill_manual(values = palette5) +noTime
    # scale_x_continuous(breaks = c(43, 44)\) +
    labs(
      title = "Mean absolute errors",
      subtitle = "by model and week",
      x = "Week"
    ) +
    plotTheme +
    theme(legend.position = "bottom")

unnest <- test_week_nest %>% unnest()

```


```{r validation: plot predicted and observed by hour}


mondays <-
  mutate(station_panel, monday = ifelse(day == "Mon", hour(interval60) == 1, 0)) %>%
  filter(monday != 0) %>%
  dplyr::select(interval60) %>%
  distinct()
  
halloween <- as.POSIXct("2021-10-31 01:00:00 UTC-4")

pred_time_plot_data <- 
  week_predictions %>%
  mutate(
    interval60 = map(data, pull, interval60),
    station = map(data, pull, station),
  ) %>%
  dplyr::select(interval60, station, Observed, Prediction, Regression) %>%
  unnest() %>%
  rename(Predicted = Prediction) %>%
  gather(variable, value, -Regression, -interval60, -station) %>%
  group_by(Regression, variable, interval60) %>%
  summarize(value = mean(value))

filter(pred_time_plot_data) %>%
  ggplot(aes(interval60, value, color = variable)) +
  geom_line() +
  geom_vline(xintercept = halloween, linetype = "dotted") +
  geom_vline(data = mondays, aes(xintercept = interval60), size = 0.25) +
  facet_wrap(~Regression, ncol = 1) +
  # scale_color_manual(values = palette2) +
  labs(title = "Predicted and observed trips by hour",
       subtitle = "Solid lines for Mondays; dotted line for Halloween",
       x = "Trips",
       y = "Date") +
  plotTheme +
  theme(legend.position = "bottom")

```



```{r cross-validation}


```

